---
title: "bulk_DGE_tutorial"
output: html_document
date: "2026-04-03"
---

## R Basics

R software is available here:
https://cran.r-project.org/ 

R Studio IDE can be downloaded here
https://posit.co/download/rstudio-desktop/

Packages/tools can be downloaded from the CRAN

Specific bioinformatics tools can be downloaded from Bioconductor
https://www.bioconductor.org/ 

When installing from the CRAN you can use:
```{r,echo=FALSE}
#install.packages("PACKAGENAME")
```

When installing from Bioconductor use:
```{r,echo=FALSE}
#BiocManager::install("PACKAGENAME")
```

Once packages are installed, you can load them using:
```{r,echo=FALSE}
#library(PACKAGENAME)
```

First let's set our working directory so we can export/import files from on our computer
Session -> Set Working Directory -> Choose Directory

```{r,echo=FALSE,message=F}
library(edgeR)
library(EnhancedVolcano)
library(tweeDEseqCountData)
library(tidyverse)
library(edgeR)
library(EnhancedVolcano)
library(openxlsx)
library(biomaRt)
```

Now let's load our example data

Calling this will give us an "ExpressionSet" Object that we will pull counts from
```{r,echo=T}
data(pickrell1)

?pickrell
```

this will store the counts as a matrix for us to work with
```{r,echo=T}
Counts <- exprs(pickrell1.eset)
```

running this or clicking "Counts" in our Environment will let us see what the matrix looks like
```{r,echo=FALSE}
head(Counts)
```

Counts is our R Object
For indexing we use square brackets []
In an R matrix or dataframe, we call: OBJECT[rows, columns]
We can call individual rows/columns or a set of rows/columns by using a colon

Running this will give us the first five rows and columns of the data and show us in our Console
```{r,echo=T}
Counts[1:5,1:5]
```
this gives us the dimensions of our matrix, we have 38,415 rows/genes and 69 columns/samples
```{r,echo=T}
dim(Counts)
```

Note that the rows are gene names that use ENSEMBL Gene IDs and Columns are samples

This example wants to look at DGE differences in sex since we have that information stored
the ExpressionSet object has metadata information that is stored/accessed using the '$'

Dataframes and some list objects can be queried using '$', which typically refers to a column
This will give us a vector that is considered a category with two levels (male/female)
```{r,echo=T}
Gender <- pickrell1.eset$gender
```

Since this is a vector (1 dimension), we can use length() to see that this matches the number of samples in our matrix
```{r,echo=T}
length(Gender)
```

Note that the "length" of this vector matches the columns of our matrix

We can also see what the first 5 observations look like
There's no need to use a comma because this is only in 1-dimension
```{r,echo=T}
Gender[1:5]
```


Note that we have "Levels", which mean this is a factor() or category that has been labeled already.
It tells us that our categories are female and male.
table() will summarize how many males and females make up this vector

```{r,echo=T}
table(Gender)
```

Now we're ready to make our DGEList object that is required for edgeR
The data must be integer counts (no fractions/proportions) and the rownames must be set as the gene names/symbols for it to run
```{r,echo=T}
#This has information regarding Ensembl IDs and their corresonding gene symbols
#We will use this to index our DGEList object, to make sure we're only analyzing genes that have annotations
data(annotEnsembl63)
annot <- annotEnsembl63[,c("Symbol","Chr")]
rm(annotEnsembl63)

#This was taken from another tutorial that used limma/voom
#We're only going to use this step to make a vector of genes that aren't missing
y <- DGEList(counts=Counts, group=Gender,genes=annot[rownames(Counts),])
hasannot <- rowSums(is.na(y$genes))==0

```

## DGE Example - DESeq2
Now that we have a sense of what the edgeR output looks like, let's compare it to DESeq2.
Let's make a key that defines which sample names belong to which group from Gender.

```{r,echo=T,message=F}
library(DESeq2)
df_key = cbind.data.frame(colnames(Counts),Gender ) 
names(df_key) = c("Sample","Gender")

head(df_key)
```

Normalizing to housekeeping genes tends to give more stable estimates

```{r,echo=T}
housekeeping_genes = c("ABCF1",
                       "ACTB",
                       "ALAS1",
                       "B2M",
                       "CLTC",
                       "G6PD",
                       "GAPDH",
                       "GUSB",
                       "HPRT1",
                       "LDHA",
                       "PGK1",
                       "POLR1B",
                       "POLR2A",
                       "RPL19",
                       "RPLP0",
                       "SDHA",
                       "TBP",
                       "TUBB")

housekeeping_ENS = rownames(annot[annot$Symbol %in% housekeeping_genes,])
```

Let's see if data separate out by our condition by using PCA
```{r, echo=T}


pca_plot = function(df,key){
  
  #df = Counts+1
  #key = df_key
  
  #making sure we're only using genes that have annotations
  df = df[hasannot,]
  
  #cleaning up the category in case there happens to be unwanted white space
  key$Gender = trimWhiteSpace(key$Gender) 
  
  #Gender analysis
  ds <- DESeqDataSetFromMatrix(countData=round(df), colData=key, design= ~Gender)
  ds$Gender <- relevel(ds$Gender,"female") #keeping our reference groups consistent
  isControl <- rownames(ds) %in% housekeeping_ENS
  ds = estimateSizeFactors(ds,controlGenes=isControl)
  dds <- DESeq(ds)
  
  #see steps here: https://bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html#pca-plot
  
  #applying modeling corrections with this
  vst <-vst(dds,blind=F)
  
  pcaData <- plotPCA(vst, intgroup = c( "Gender"),pcsToUse = c(1,2) ,returnData = TRUE,ntop=50)
  percentVar <- round(100 * attr(pcaData, "percentVar"))
  
  #this seems to separate CAR and cycle variation
  pca_gender_pc1_pc2 = ggplot(pcaData, aes(x = PC1, y = PC2, color = Gender)) +
    geom_point(size =3) +
    scale_color_manual(values=c("#0000ff","#00ff00"),name="Sex") +
    xlab(paste0("PC1: ", percentVar[1], "% variance")) +
    ylab(paste0("PC2: ", percentVar[2], "% variance")) +
    coord_fixed() +
    theme_bw()
  
  #sometimes other PCs might give better separation
  pcaData <- plotPCA(vst, intgroup = c( "Gender"),pcsToUse = c(1,3) ,returnData = TRUE,ntop=50)
  percentVar <- round(100 * attr(pcaData, "percentVar"))
  
  #this seems to separate CAR and cycle variation
  pca_gender_pc1_pc3 = ggplot(pcaData, aes(x = PC1, y = PC3, color = Gender)) +
    geom_point(size =3) +
    scale_color_manual(values=c("#0000ff","#00ff00"),name="Sex") +
    xlab(paste0("PC1: ", percentVar[1], "% variance")) +
    ylab(paste0("PC3: ", percentVar[2], "% variance")) +
    coord_fixed() +
    theme_bw()
  
  pcaData <- plotPCA(vst, intgroup = c( "Gender"),pcsToUse = c(1,4) ,returnData = TRUE,ntop=50)
  percentVar <- round(100 * attr(pcaData, "percentVar"))
  
  #this seems to separate CAR and cycle variation
  pca_gender_pc1_pc4 = ggplot(pcaData, aes(x = PC1, y = PC4, color = Gender)) +
    geom_point(size =3) +
    scale_color_manual(values=c("#0000ff","#00ff00"),name="Sex") +
    xlab(paste0("PC1: ", percentVar[1], "% variance")) +
    ylab(paste0("PC4: ", percentVar[2], "% variance")) +
    coord_fixed() +
    theme_bw()
  
  return(list("pca_gender_pc1_pc2" = pca_gender_pc1_pc2,
              "pca_gender_pc1_pc3" = pca_gender_pc1_pc3,
              "pca_gender_pc1_pc4" = pca_gender_pc1_pc4
              
  ))
  
  
}

pca_figures = pca_plot(Counts,df_key)

pca_figures$pca_gender_pc1_pc2
pca_figures$pca_gender_pc1_pc3
pca_figures$pca_gender_pc1_pc4
```

There doesn't seem to be great separation, however, PCAs are just descriptive representations. This doesn't mean there aren't relevant DEGs. PCA is NOT a hypothesis test! Try not to over-interpret anything from a PCA.



Now we can export whichever PCA plot we want.
```{r}
ggsave(
  "PCA_PC1_PC2.png",
  pca_figures$pca_gender_pc1_pc2,
  width = 10,
  height = 10,
  dpi = 300,
  device="png"
)
```



```{r,echo=T}
dge_deseq2 = function(df,key){
  #df = Counts
  #key = df_key
  
  #making sure we're only using genes that have annotations
  df = df[hasannot,]
  
  #cleaning up the category in case there happens to be unwanted white space
  key$Gender = trimWhiteSpace(key$Gender) 
  
  #Gender analysis
  ds <- DESeqDataSetFromMatrix(countData=round(df), colData=key, design= ~Gender)
  ds$Gender <- relevel(ds$Gender,"female") #keeping our reference groups consistent
  isControl <- rownames(ds) %in% housekeeping_ENS
  ds = estimateSizeFactors(ds,controlGenes=isControl)
  ds <- DESeq(ds)
  #resultsNames(ds)
  dge_gender = as.data.frame(results(ds, contrast=list(c("Gender_male_vs_female")), test="Wald",pAdjustMethod = "BH"))
  
  #merge this with the count table like we did before
  
  dge_gender_full = transform(merge(dge_gender,annot,by="row.names"),row.names=Row.names,Row.names=NULL)
  
  
  results = list("Gender_DESeq2" = dge_gender_full,
                     "counts" = df[order(Gender),]
  )
  
  return(results)
  
  
  
}

deseq2_output = dge_deseq2(Counts+1,df_key)
```

```{r,echo=T}
deseq2_volcano = EnhancedVolcano(deseq2_output$Gender_DESeq2,
                lab = deseq2_output$Gender_DESeq2$Symbol,
                x = 'log2FoldChange',
                y = 'padj',
                pCutoff = 0.05,
                FCcutoff = .5,
                pointSize = 1.0,
                labSize = 3.0,
                legendLabels=c('NS','Log2FC','p-value',
                               'p-value & Log2FC'),
                legendPosition = 'none',
                legendLabSize = 10,
                legendIconSize = 1,
                axisLabSize = 10,
                title = "",
                subtitle = "",
                caption = "",
                titleLabSize = 0,
                subtitleLabSize = 0,
                captionLabSize = 0) 

deseq2_volcano
```

Unsurprisingly, the right side of the plot is a Y chrom associated gene annotation the left side is an X chrom associated gene annotation. Since we defined our reference as "female" this means that the up-regulated genes are those that are higher in males compared to females.


Export output of interest
```{r,echo=T}
write.xlsx(deseq2_output$Gender_DESeq2,"DESeq2_output.xlsx")
```

```{r,echo=T}
ggsave(
  "volcano_DESeq2.png",
  deseq2_volcano,
  width = 10,
  height = 10,
  dpi = 300,
  device="png"
)
```

